Supporting Non-Functional Requirements in Services Software Development Process: An MDD Approach

### 2013-10-21 SOFSEM2014/FinalVersion/introduction.tex
12 pages!!

Model Driven Development (MDD)~\cite{Favre06arigorous} is a top-down approach for  designing and developing software systems 
%The main ideas of MDD were originaly 
proposed by the Object Management Group
(OMG)\footnote{\texttt{http://www.omg.org/mda}.}. 
MDD provides a set of
guidelines for  structuring  specifications using
 \textit{models} that specify a software system at different levels of abstraction or \textit{viewpoints}: 
%MDD defines three viewpoints:
%
\begin{trivlist}
\item \textit{Computation Independent Models (CIM):} This viewpoint represents the software system at its highest level of abstraction. It focusses on the system
environment, and on its business and requirement specifications. 
At this moment of the development, the structure and system processing details are still unknown or undetermined. 
\item \textit{Platform Independent Models (PIM):} This viewpoint focusses on the system functionality, hiding the details of any particular platform. 
%The specification defines those parts of the system that are independent of the development platform. 
\item \textit{Platform Specific Models (PSM):} This viewpoint focusses on the functionality, in the context of a particular implementation platform.
Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.  
\end{trivlist}
%
Besides the notion of model at each level of abstraction, MDD requires the use of \textit{model transformations} between levels.
These transformations may be automatic or semi-automatic and implement the refinement process between levels.

%\bigskip 
MDD has been applied for developing service oriented applications.
In Service-Oriented Computing~\cite{Papazoglou2007}, pre-existing services are
combined to produce applications and provide the business logic. The selection of services is usually guided by the functional requirements of the application being developed. 
Some methodologies and techniques have been proposed to help the software
developer in the specification of functional requirements of the business logic,
such as the Service-Oriented Development Method
(SOD-M)~\cite{decastro1}. 
%SOD-M is based on MDD and proposes
%models, practices and techniques to aid in software development. 
%SOD-M does not
 %suport  the specification of non-functional requirements, such as
%security, reliability, and efficiency.

Ideally, non-functional requirements such as
security, reliability, and efficiency
would be considered along with all the stages of the software development. The
adoption of non-functional specifications from the early states of development
can help the developer to produce applications that are capable of dealing with
the application context.
%
%\bigskip
%
Non-functional properties of service-oriented applications  have been
addressed in academic works and standards~\cite{ws-co,ws-tra,wsci}.
Dealing with these kind of properties involves the use of specific technologies
in different layers of the SOC architecture, for instance during the description
of service APIs (such as WSDL\cite{wsdl} or REST~\cite{rest}) or to express
service coordinations (like WS-BPEL~\cite{bpel03}).

>>>>

Model Driven Development (MDD)~\cite{Favre06arigorous} is a top-down approach for  designing and developing software systems 
%The main ideas of MDD were originaly 
proposed by the Object Management Group
(OMG)\footnote{\texttt{http://www.omg.org/mda}.}. 
MDD provides a set of
guidelines for  structuring  specifications using
 \textit{models} that specify a software system at different levels of abstraction or \textit{viewpoints}: 
%MDD defines three viewpoints:

\noindent \textit{Computation Independent Models (CIM):} This viewpoint represents the software system at its highest level of abstraction. It focusses on the system
environment, and on its business and requirement specifications. 
At this moment of the development, the structure and system processing details are still unknown or undetermined.

\noindent \textit{Platform Independent Models (PIM):} This viewpoint focusses on the system functionality, hiding the details of any particular platform. 
%The specification defines those parts of the system that are independent of the development platform.

\noindent \textit{Platform Specific Models (PSM):} This viewpoint focusses on the functionality, in the context of a particular implementation platform.
Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system.

Besides the notion of model at each level of abstraction, MDD requires the use of \textit{model transformations} between levels.
These transformations may be automatic or semi-automatic and implement the refinement process between levels.

%\bigskip 
MDD has been applied for developing service oriented applications.
In Service-Oriented Computing~\cite{scube2010book}, pre-existing services are
combined to produce applications and provide the business logic. The selection of services is usually guided by the functional requirements of the application being developed. 
Some methodologies and techniques have been proposed to help the software
developer in the specification of functional requirements of the business logic,
such as the Service-Oriented Development Method
(SOD-M)~\cite{decastro1}. 
%SOD-M is based on MDD and proposes
%models, practices and techniques to aid in software development. 
%SOD-M does not
 %suport  the specification of non-functional requirements, such as
%security, reliability, and efficiency.

Ideally, non-functional requirements such as
security, reliability, and efficiency
would be considered along with all the stages of the software development. The
adoption of non-functional specifications from the early states of development
can help the developer to produce applications that are capable of dealing with
the application context.
%
%\bigskip
%
Non-functional properties of service-oriented applications  have been
addressed in academic works and standards.%~\cite{ws-co,ws-tra,wsci}.
Dealing with these kind of properties involves the use of specific technologies
in different layers of the SOC architecture, for instance during the description
of service APIs (such as WSDL\cite{wsdl} or REST~\cite{rest}) or to express
service coordinations (like WS-BPEL~\cite{bpel03}).

### 2013-10-21 SOFSEM2014/FinalVersion/main.tex
12 pages!!

\author{Valeria de Castro\inst{1} \and Martin A. Musicante\inst{2} \and Umberto Souza da Costa\inst{2} \and Plácido~A.~de~Souza~Neto\inst{3} \and Genoveva Vargas-Solar\inst{4}}

@@

\maketitle
\begin{abstract}
This paper presents the $\pi$-SODM method an extension to the Service Oriented Development Method (SOD-M) to support the developement of services software considering their functional and    non-functional requirements. 
Specifically, $\pi$-SODM proposes: 
\textit{(i)} meta-models for representing non-functional requirements in different abstraction levels; 
\textit{(ii)} model-to-model transformation rules, useful to semi-automatically refine Platform Independent Models into Platform Specific Models; and 
\textit{(iii)} rules to transform Platform Specific Models into concrete implementations. 
In order to illustrate the use of this methodology the paper describes how its use to develop a proof-of-concept.
\end{abstract}

@@

\section{Related work}
\label{sec:related}
Over the last years, a number of approaches have been proposed for the
development of web services. These approaches range from the proposal of new 
languages for web service descriptions~\cite{bpel03,SBS04} 
to techniques to support phases of the development cycle of this kind of
software~\cite{lipari2007}. In general, these
approaches concentrate on specific problems, like supporting transactions or
QoS, in order to improve the security and reliability of service-based
applications. Some proposals address service composition: workflow
definition~\cite{AalstHKB03,MuP06} or semantic equivalence between
services~\cite{BHM06}. The proposed solutions come from many communities, 
including those of Theoretical Computer Science~\cite{VA05,GGP08}, 
Software Engineering~\cite{Aal03,MendesPDB09}, Programming
Languages~\cite{MPC08,bpel03} and Databases~\cite{ABM01}.

Despite the variety of techniques proposed, there is not yet a consensus on a software
process methodology for web services. Some methodologies address the service-based 
development towards a standard or a new way to develop reliable applications. 
SOD-M and SOMF \cite{somf} are MDD approaches for web services; 
S-Cube \cite{scube2010book} is focused on the representation of business processes and 
service-based development; SOMA \cite{soma} is a methodology described by IBM
for SOA solutions; Extended SOA \cite{PapazoglouH06} merges RUP and
BPM\cite{bpm} concepts for service modeling; DEVISE \cite{DEVISE} is a
methodology for building service-based infrastructure for collaborative
enterprises. Other proposals include, the WIED model \cite{TongrungrojanaL04}, 
that acts as a bridge between business modeling and design models, and traditional 
approaches for software engineering \cite{sommerville08} applied to SOC.

>>>>

\author{Valeria de Castro\inst{1} \and Martin A. Musicante\inst{2} \and Umberto Souza da Costa\inst{2} \and Pl\'acido~A.~de~Souza~Neto\inst{3} \and Genoveva Vargas-Solar\inst{4}}

@@

\maketitle
\begin{abstract}
This paper presents the $\pi$-SODM method an extension to the Service Oriented Development Method (SOD-M) to support the development of services software considering their functional and    non-functional requirements. 
Specifically, $\pi$-SODM proposes: 
\textit{(i)} meta-models for representing non-functional requirements in different abstraction levels; 
\textit{(ii)} model-to-model transformation rules, useful to semi-automatically refine Platform Independent Models into Platform Specific Models; and 
\textit{(iii)} rules to transform Platform Specific Models into concrete implementations. 
In order to illustrate the use of this methodology the paper describes how its use to develop a proof-of-concept.
\end{abstract}

@@

\section{Related work}
\label{sec:related}
Over the last years, a number of approaches have been proposed for the
development of web services. 
These approaches range from the proposal of new 
languages for web service descriptions~\cite{bpel03,SBS04} 
to techniques to support phases of the development cycle of this kind of
software~\cite{lipari2007}. 
In general, these approaches concentrate on specific problems, like supporting transactions or
QoS, in order to improve the security and reliability of service-based
applications. 
Some proposals address service composition: workflow
definition~\cite{AalstHKB03,MuP06} or semantic equivalence between
services~\cite{BHM06}. 
%The proposed solutions come from many communities, 
%including those of Theoretical Computer Science~\cite{VA05,GGP08}, 
%Software Engineering~\cite{Aal03,MendesPDB09}, Programming
%Languages~\cite{MPC08,bpel03} and Databases~\cite{ABM01}.

Works dealing with non-functional properties in service-oriented 
development can be organized in two main groups: those working on the \textit{modeling 
of particular non-functional properties or QoS attributes} and those 
proposing \textit{architectures or frameworks to manage and 
validate QoS attributes in web service composition processes}. 
The first group considers specific non-functional concerns (e.g. security)
which is modelled and then associated to functional models of the application. 
The work of Chollet et al.~\cite{CholletL09} defines a proposal to associate non-functional 
quality properties (security properties) to functional activities in a web service composition model. Schmeling et al.~\cite{SchmelingCM11} present 
an approach and also a toolset for specifying and implementing non-functional 
concerns in web service compositions.
Non-functional concerns are modelled and then related to a service composition 
represented in a BPMN diagram. 
Ovaska et al.~\cite{OvaskaEHPA10} present an approach to 
support quality management at design time. 
Quality requirements are modelled in a first phase and then
represented in an architectural model where quality requirements are associated to some
components of the model.

%We propose to model non-functional properties together with the modelling of functional concerns.
%We claim what this approach simplifies the web service development process. 
%Non-functional properties are represented in our work as constraints and policies 
%but more general quality model such as the proposed by~\cite{Goeb2011,Klass2009} 
%could be taken in consideration in further works.

The second group of works dealing with non-functional requirements for services 
propose specific architectures or frameworks to manage and validate 
QoS attributes in service composition processes~\cite{XiaoCZBOLH08,Babamir2010,Karunamurthy2012787}. 
%Although our proposal focus on the design of service oriented application with non-functional 
%properties and not on the validation of QoS attributes at runtime.

Despite the variety of techniques proposed, there is not yet a consensus on a software methodology 
for web services. 
Some methodologies address the service-based 
development towards a standard or a new way to develop reliable applications. 
SOD-M and SOMF~\cite{somf} are MDD approaches for web services; 
S-Cube~\cite{scube2010book} is focused on the representation of business processes and 
service-based development; SOMA~\cite{soma} is a methodology 
for SOA solutions; 
%Extended SOA~\cite{PapazoglouH06} merges RUP and
%BPM~\cite{bpm} concepts for service modelling; 
DEVISE~\cite{DEVISE} is a
methodology for building service-based infrastructure for collaborative
enterprises. 
Other proposals include, the WIED model~\cite{TongrungrojanaL04}, 
that acts as a bridge between business modeling and design models, and traditional 
approaches for software engineering applied to SOC.

### 2013-10-21 SOFSEM2014/FinalVersion/pisodm.tex
12 pages!!

\begin{figure}[h]
\centering
\includegraphics[width=1.0\textwidth]{figs/piSODM}
\caption{$\pi$SOD-M.}
\label{fig:piSOD-M}
\end{figure}

The \textit{$\pi$-UseCase} meta-model describes functional and non-functional requirements.
Non-functional requirements are defined as \textit{constraints} over processing and data.

The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent restrictions over data and actions that must be performed upon certain conditions.

The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
described in the \textit{$\pi$-UseCase} model into contracts that are associated
with services.

The \textit{$\pi$-ServiceComposition} meta-model provides the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a}
which put together contracts with similar non-functional requirements. 
For instance, security and privacy restrictions may be grouped into a security policy.
\textit{$\pi$-ServiceComposition} models can be refined into PSMs. Policies are associated to service operations and combine \textit{constraints} and \textit{reactive recovery actions}.
Constraints are restrictions that must be verified during the execution of the application. 
Failure to verify the constraints will trigger exceptions to execute their corresponding recovery actions.
An example of policy is the requirement of authentication for executing some of the system functions. 
The action associated to this policy may perform the authentication of the user.

The \textit{$\pi$-PEWS} meta-model is a PSM (see Figure~\ref{fig:piSOD-M}).
At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
The \textit{$\pi$-PEWS} meta-model is the PSM adopted in this work.
\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into PEWS~\cite{BaCAM05,Placido2010LTPD} or BPEL~\cite{bpel03} code.
Although PEWS is our language of choice, other composition languages can be used as target.
%This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.

@@

%
%Similarly to SOD-M, our approach targets the construction of service-oriented applications that implement business processes.
Thus, $\pi$SOD-M proposes a development process based on the definition of models
(instances of the meta-models) and transformations between models.
There are two kinds of transformations:
Model-to-model transformations are used during the software process to refine the specification.
Model-to-text transformations are the last step of the process and generate code.
In the next section we develop an example, to serve as a proof-of-concept.
The example will show the actual notation used for models.

>>>>

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth]{figs/piSOD-M.png}
\caption{$\pi$SOD-M.}
\label{fig:piSOD-M}
\end{figure}
%
The \textit{$\pi$-UseCase} meta-model describes functional and non-functional requirements.
Non-functional requirements are defined as \textit{constraints} over processing and data. 
%
The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent restrictions over data and actions that must be performed upon certain conditions. 
%
The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
described in the \textit{$\pi$-UseCase} model into contracts that are associated
with services. 
%
The \textit{$\pi$-ServiceComposition} meta-model provides the concept of \textit{Policy}~\cite{Espinosa-Oviedo2011a}
which put together contracts with similar non-functional requirements. 
For instance, security and privacy restrictions may be grouped into a security policy.
\textit{$\pi$-ServiceComposition} models can be refined into PSMs. Policies are associated to service operations and combine \textit{constraints} and \textit{reactive recovery actions}.
Constraints are restrictions that must be verified during the execution of the application. 
Failure to verify the constraints will trigger exceptions to execute their corresponding recovery actions.
An example of policy is the requirement of authentication for executing some of the system functions. 
The action associated to this policy may perform the authentication of the user.
%
The \textit{$\pi$-PEWS} meta-model is a PSM (see Figure~\ref{fig:piSOD-M}).
At the PSM level we have lower-level models that can be automatically translated into actual computer programs.
The \textit{$\pi$-PEWS} meta-model is the PSM adopted in this work.
\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into PEWS~\cite{BHM06} %~\cite{BaCAM05,Placido2010LTPD} 
or BPEL~\cite{bpel03} code.
Although PEWS is our language of choice, other composition languages can be used as target.
%This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.

@@

%
%Similarly to SOD-M, our approach targets the construction of service-oriented applications that implement business processes.
Thus, $\pi$SOD-M proposes a development process based on the definition of models
(instances of the meta-models) and transformations between models.
There are two kinds of transformations:
Model-to-model transformations are used during the software process to refine the specification.
Model-to-text transformations are the last step of the process and generate code.

%%%BEGIN{Text Included by Placido Neto for Camera-Ready}%%%%%%
$\pi$SOD-M environment is built on the top of Eclipse. 
We also used the Eclipse Modelling Framework (EMF) to define, edit and handle
(meta)-models. 
To automate the transformation models we use ATL~\cite{atl_manual} and Acceleo~\cite{acceleo}.

In the next section we develop an example, to serve as a proof-of-concept.
The example will show the actual notation used for models.

### 2013-10-21 SOFSEM2014/FinalVersion/proofOfConcept.tex
12 pages!!

Non-functional requirements are specified by rules and conditions to be verified during the execution of tasks.
In our example we have the following non-functional requirements:
\begin{numtrivlist}
\item Twitter requires  an authentication protocol that allows three login failures before blocking. 
\item Crime notification needs privileged access.
\item Civilian users can only track crimes for which they have clearance: Civilian population cannot track all the crimes notified by the police. 
\item If \textsf{Google Maps} is unavailable, the results are delivered as text. 
\item Querying about crimes without having proper clearance yields an empty map.
\item Access rights to detailed information depends on user clearance and zone assignment for police officers. 
\item The application maintains a detailed log. 
\end{numtrivlist}
The idea about these requirements is to leave the application logic expressed by functional requirements as independent as possible from exceptional situations like the unavailability of a service and the conditions in which services are called for example through an authentication protocol. These requirements can be  weaved as activities and branches of the composition or kept independent. The interest of the second option is that the maintenance and the evolution of the application logic can be easier. For example, the services called by the application are not hard coded (Twitter and Google Maps in the example), neither the actions to deal with exceptions (replacing anohter Map service or doing nothing).

@@

\paragraph*{\textit{$\pi$-UseCase} model:} 
In our example we have five use cases (Figure~\ref{fig:piUC}), which represent the
system functions (tasks) and constraints. 
We will not detail the functional part of the specification, due to lack of space.
The constrains defined for our tracking crime example are:

\begin{trivlist}
  \item[-] The \textsf{Notify crime} task requires that the user is logged in. 
  This is an example of a \textit{value constraint}, where the value associated to the condition depends on the semantics of the application.
  In this case, it represents the maximum number of allowed login attempts;
\item[-] The \textsf{Store crimes} task requires the verification of the user's clearance (also a value constraint). 
\item[-] In order to perform the \textsf{Track crimes} task, it is necessary that the notifier user is in the contact list of the requesting user.
This is an example of \textit{business constraint}.
Additionally the requesting user must be logged in.
\item[-]  For the \textsf{View Crime Map} task, the specification defines that if the Google Maps service is not available, the result is presented as text. This is an example of \textit{exceptional behaviour constraint}. 
The availability of the Google Maps service is verified by a \textit{business constraint}.
 \item[-] The \textsf{Show crime details} task is specified to have three constraints: A \textit{value constraint} is defined to verify the user's clearance level; A \textit{business constraint} is used to ensure that the user's clearance is valid for the geographic zone of the crime; Another \textit{value constraint} defines that the log is to be maintained.
\end{trivlist}
\begin{figure}[ht!]
\centering
\includegraphics[width=0.8\textwidth]{figs/piUseCase}
\caption{$\pi$-UseCase Model}
\label{fig:piUC}
\end{figure}
\paragraph*{\textit{$\pi$-ServiceProcess} model:}
The model presented in Figure~\ref{fig:piUC} is transformed, at this stage of the development, into a similar graph, where \textit{(i)} the task nodes are better detailed, by refining the control and data flows; and \textit{(ii)} constraints are transformed into \textit{contracts} (pre- and post-conditions).
The new model describes the application's activities and defines contracts for each activity or for parts of the application.

@@

\begin{figure}[ht!]
\centering
\includegraphics[width=0.99\textwidth]{figs/piServiceProcess}
\caption{$\pi$-\textit{ServiceProcess} Model}
\label{fig:piSP}
\end{figure}

@@

\begin{figure}[t]
\centering
\subfloat[$\pi$-ServiceComposition Model]
{\label{fig:piSC}\includegraphics[width=0.95\textwidth]{figs/piServiceComposition}}
~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
\\
\subfloat[$\pi$-ServiceComposition Policies]
{\label{fig:crimePolicies}\includegraphics[width=0.85\textwidth]{figs/policies}}
~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
\caption{Service Composition and Policies.}
\label{fig:policies}
\end{figure}
This model describes the service compositions and their related business
collaborators (external services), policies, rules associated with a policy and
the whole application process and functionalities.
The workflow execution will consider the composition of business services and
their specific Business Collaborators. 
Each action will trigger the execution of one or more external services. 
Before and after each service call, policy rules are verified.

@@

\paragraph*{$\pi$\textit{-PEWS} Model:}
These models are produced by a model-to-text transformation that takes a $\pi$-\textit{ServiceComposition} model and generates $\pi$-\textit{PEWS} specification code.
This code is a service composition program that can be compiled into executable code.
$\pi$-\textit{PEWS} models are expressed in a variant of the PEWS composition language.

\begin{figure}[t]
\begin{footnotesize}
%\begin{lstlisting}[label=list:rulePre,caption=ATL - piServiceComposition2piPEWS : Pre-condition Rule. ]
%\begin{lstlisting}[label=list:rulePre]
\begin{verbatim}
//Namespaces specify service URI
1 namespace twitter = www.twitter.com/service.wsdl
2 namespace googlemaps = maps.googleapis.com/maps/api/service
3 namespace amazondynamodb = rds.amazonaws.com/doc/2010-07-28/AmazonRDSv4.wsdl
4 namespace police = www.police.fr/service.wsdl
//Operations 
5 alias publishTwitter = portType/publishTwitter in twitter
6 alias searchCrime = portType/searchCrime in amazondynamodb 
7 alias showMap = portType/showMap in googlemaps 
...
//Services
8  service notifyCrime = publishCrime  . publishTwitter 
9  service trackCrime= searchCrime . verifyService
10 Service visualizeCrime =  showMap . getCrimeDetail
...
//Path
11 (notifyCrime.storeiCrime) || (trackCrime.visualizeCrime.getCrimeDetail)
...
//Contracts
12 defContract notifyCrimeContract{ 
13    isAppliedTo: notifyCrime
14	  requires: userId == ?? && passw == ?? && req(notifyCrime)  < 3
15 	      (OnFailureDo: NOT(action_publish(crime));
16	  ensures: publishTwitter(crime) == true
17 	      (OnFailureDo: skip);
28 }
\end{verbatim}
% \end{lstlisting}
\end{footnotesize}
\caption{$\pi$-PEWS code for the crime tracking example (partial, simplified).\label{fig:pewscontract} }
\end{figure}

The $\pi$-PEWS program generated from the model in Figure~\ref{fig:policies} is partially presented in Figure~\ref{fig:pewscontract}. 
The figure shows a simplified program code, produced in accordance to the following guidelines:
\begin{numtrivlist}
\item Namespaces, identifying the addresses of external services are produced from the Business Collaborators of the higher-level model. 
We define four of them, corresponding to the Police, Twitter, Google Map and Amazon partners.
\item Specific operations exported by each business collaborator are identified to an \textit{operation} of the program (Each operation is given an \texttt{alias}).
\item The workflow in Figure~\ref{fig:piSC} is translated into the text in line 11.
\item \textit{Contracts} are defined in $\pi$-PEWS as having pre-conditions (\texttt{requires}), post-conditions (\texttt{ensures}) and actions (\texttt{OnFailureDo}) to be executed case a condition is not verified. 
Contracts are generated from Policies (such as those of Figure~\ref{fig:piSC}.
\end{numtrivlist}

>>>>

Non-functional requirements are specified by rules and conditions to be verified during the execution of tasks.
In our example we have the following non-functional requirements:
\begin{numtrivlist}
\item Twitter requires authentication and allows three login failures before blocking. 
\item Crime notification needs privileged access.
\item Civilian users can only track crimes for which they have clearance: Civilian population cannot track all the crimes notified by the police. 
\item If \textsf{Google Maps} is unavailable, the results are delivered as text. 
\item Querying about crimes without having proper clearance yields an empty map.
\item Access rights to detailed information depends on user clearance and zone assignment for police officers. 
\item The application maintains a detailed log. 
\end{numtrivlist}
The idea about these requirements is to leave the application logic expressed by functional requirements as independent as possible from exceptional situations like the unavailability of a service and the conditions in which services are called for example through an authentication protocol. These requirements can be  weaved as activities and branches of the composition or kept independent. The interest of the second option is that the maintenance and the evolution of the application logic can be easier. For example, the services called by the application are not hard coded (Twitter and Google Maps in the example), neither the actions to deal with exceptions (replacing anohter Map service or doing nothing).

@@

\paragraph*{\textit{$\pi$-UseCase} model:} 
In our example we have five use cases (Figure~\ref{fig:piUC}), which represent the
system functions (tasks) and constraints. 
We will not detail the functional part of the specification, due to lack of space.
The constrains defined for our tracking crime example are: 
\begin{trivlist}
  \item[-] The \textsf{Notify crime} task requires that the user is logged in. 
  This is an example of a \textit{value constraint}, where the value associated to the condition depends on the semantics of the application.
  In this case, it represents the maximum number of allowed login attempts;
\item[-] The \textsf{Store crimes} task requires the verification of the user's clearance (also a value constraint). 
\item[-] In order to perform the \textsf{Track crimes} task, it is necessary that the notifier user is in the contact list of the requesting user.
This is an example of \textit{business constraint}.
Additionally the requesting user must be logged in.
\item[-]  For the \textsf{View Crime Map} task, the specification defines that if the Google Maps service is not available, the result is presented as text. This is an example of \textit{exceptional behaviour constraint}. 
The availability of the Google Maps service is verified by a \textit{business constraint}.
 \item[-] The \textsf{Show crime details} task is specified to have three constraints: A \textit{value constraint} is defined to verify the user's clearance level; A \textit{business constraint} is used to ensure that the user's clearance is valid for the geographic zone of the crime; Another \textit{value constraint} defines that the log is to be maintained.
\end{trivlist}
\begin{figure}[t]
\centering
\includegraphics[width=0.8\textwidth]{figs/piUseCase}
\caption{$\pi$-UseCase Model}
\label{fig:piUC}
\end{figure}
\paragraph*{\textit{$\pi$-ServiceProcess} model:}
The model presented in Figure~\ref{fig:piUC} is transformed, at this stage of the development, into a similar graph, where \textit{(i)} the task nodes are better detailed, by refining the control and data flows; and \textit{(ii)} constraints are transformed into \textit{contracts} (pre- and post-conditions).
The new model describes the application's activities and defines contracts for each activity or for parts of the application.

@@

\begin{figure}[t]
\centering
\includegraphics[width=1.05\textwidth]{figs/piServiceProcess}
\caption{$\pi$-\textit{ServiceProcess} Model}
\label{fig:piSP}
\end{figure}

@@

\begin{figure}[t]
\centering
\subfloat[$\pi$-ServiceComposition Model]
{\label{fig:piSC}\includegraphics[width=0.95\textwidth]{figs/piServiceComposition}}
~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
\\[-1mm]
\subfloat[$\pi$-ServiceComposition Policies]
{\label{fig:crimePolicies}\includegraphics[width=0.85\textwidth]{figs/policies}}
~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. (or a blank line to force the subfig onto a new line)
\caption{Service Composition and Policies.}
\label{fig:policies}
\end{figure}
%This model describes the service compositions and their related business
%collaborators (external services), policies, rules associated with a policy and
%the whole application process and functionalities.
%The workflow execution will consider the composition of business services and
%their specific Business Collaborators. 
%Each action will trigger the execution of one or more external services. 
%Before and after each service call, policy rules are verified.

@@

\paragraph*{$\pi$\textit{-PEWS} Model:}
These models are produced by a model-to-text transformation that takes a $\pi$-\textit{ServiceComposition} model and generates $\pi$-\textit{PEWS} specification code.
This code is a service composition program that can be compiled into executable code.
$\pi$-\textit{PEWS} models are expressed in a variant of the PEWS composition language.
\begin{figure}[t]
\begin{small}
%\begin{lstlisting}[label=list:rulePre,caption=ATL - piServiceComposition2piPEWS : Pre-condition Rule. ]
%\begin{lstlisting}[label=list:rulePre]
\begin{verbatim}
//Namespaces specify service URI
1 namespace twitter = www.twitter.com/service.wsdl
2 namespace googlemaps = maps.googleapis.com/maps/api/service
3 namespace amazondynamodb = rds.amazonaws.com/doc/2010-07-28/AmazonRDSv4.wsdl
4 namespace police = www.police.fr/service.wsdl
//Operations 
5 alias publishTwitter = portType/publishTwitter in twitter
6 alias searchCrime = portType/searchCrime in amazondynamodb 
7 alias showMap = portType/showMap in googlemaps 
//Services
8  service notifyCrime = publishCrime  . publishTwitter 
9  service trackCrime= searchCrime . verifyService
10 Service visualizeCrime =  showMap . getCrimeDetail
//Path
11 (notifyCrime.storeiCrime) || (trackCrime.visualizeCrime.getCrimeDetail)
//Contracts
12 defContract notifyCrimeContract{  isAppliedTo: notifyCrime
13	  requires: userId == ?? && passw == ?? && req(notifyCrime)  < 3
14 	      (OnFailureDo: NOT(action_publish(crime));
15	  ensures: publishTwitter(crime) == true  (OnFailureDo: skip); }
\end{verbatim}
% \end{lstlisting}
\end{small}
\caption{$\pi$-PEWS code for the crime tracking example (partial, simplified).\label{fig:pewscontract} }
\end{figure}
The $\pi$-PEWS program generated from the model in Figure~\ref{fig:policies} is partially presented in Figure~\ref{fig:pewscontract}. 
The figure shows a simplified program code, produced in accordance to the following guidelines:
\begin{numtrivlist}
\item Namespaces, identifying the addresses of external services are produced from the Business Collaborators of the higher-level model. 
We define four of them, corresponding to the Police, Twitter, Google Map and Amazon partners.
\item Specific operations exported by each business collaborator are identified to an \textit{operation} of the program (Each operation is given an \texttt{alias}).
\item The workflow in Figure~\ref{fig:piSC} is translated into the text in line 11.
\item \textit{Contracts} are defined in $\pi$-PEWS as having pre-conditions (\texttt{requires}), post-conditions (\texttt{ensures}) and actions (\texttt{OnFailureDo}) to be executed case a condition is not verified. 
Contracts are generated from Policies (such as those of Figure~\ref{fig:piSC}.
\end{numtrivlist}
